---
title: "Lab 4 Homework"
author: "Sidney Parel"
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Instructions
Answer the following questions and complete the exercises in RMarkdown. Please embed all of your code and push your final work to your repository. Your final lab report should be organized, clean, and run free from errors. Remember, you must remove the `#` for the included code chunks to run. Be sure to add your name to the author header above.  

Make sure to use the formatting conventions of RMarkdown to make your report neat and clean!  

## Load the tidyverse
```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

## Data
For the homework, we will use data about vertebrate home range sizes. The data are in the class folder, but the reference is below.  

**Database of vertebrate home range sizes.**  
Reference: Tamburello N, Cote IM, Dulvy NK (2015) Energy and the scaling of animal space use. The American Naturalist 186(2):196-211. http://dx.doi.org/10.1086/682070.  
Data: http://datadryad.org/resource/doi:10.5061/dryad.q5j65/1  

**1. Load the data into a new object called `homerange`.**
```{r}
homerange <- read_csv("C:\\Users\\Owner\\Documents\\GitHub\\BIS15W2022_sparel\\lab4\\data\\Tamburelloetal_HomeRangeDatabase.csv")
```

**2. Explore the data. Show the dimensions, column names, classes for each variable, and a statistical summary. Keep these as separate code chunks.**  
The `homerange` data set has 569 rows and 24 columns.
```{r}
homerange %>% dim()
```
The column names in `homerange` are reported below.
```{r}
homerange %>% names()
```
The classes for each variable in `homerange` are reported below.
```{r}
homerange %>% spec()
```
A statistical summary for `homerange` is reported below.
```{r}
homerange %>% summary()
```

**3. Change the class of the variables `taxon` and `order` to factors and display their levels.**  
```{r}
homerange$taxon <- homerange$taxon %>% as.factor()
homerange$taxon %>% levels()
```
```{r}
homerange$order <- homerange$order %>% as.factor()
homerange$order %>% levels()
```

**4. What taxa are represented in the `homerange` data frame? Make a new data frame `taxa` that is restricted to taxon, common name, class, order, family, genus, species.** 

Lake fishes, river fishes, marine fishes, birds, mammals, lizards, snakes, turtles, and tortoises are represented in the `homerange` data frame.
```{r}
taxa <- homerange %>% select(taxon, common.name, class, order, family, genus, species)
taxa$taxon %>% unique()
```


**5. The variable `taxon` identifies the large, common name groups of the species represented in `homerange`. Make a table the shows the counts for each of these `taxon`.**  
```{r}
homerange$taxon %>% table()
```


**6. The species in `homerange` are also classified into trophic guilds. How many species are represented in each trophic guild.**  
The trophic guilds in `homerange` are "carnivore" and "herbivore".
```{r}
homerange$trophic.guild %>% unique()
```
342 species are represented in the carnivore trophic guild and 227 species are represented in the herbivore trophic guild.
```{r}
homerange %>% filter(trophic.guild == "carnivore") %>% nrow()
homerange %>% filter(trophic.guild == "herbivore") %>% nrow()
```


**7. Make two new data frames, one which is restricted to carnivores and another that is restricted to herbivores.**  
```{r}
carnivores <- homerange %>% filter(trophic.guild == "carnivore")
herbivores <- homerange %>% filter(trophic.guild == "herbivore")
```


**8. Do herbivores or carnivores have, on average, a larger `mean.hra.m2`? Remove any NAs from the data.**  
The data frames do not contain NAs in `mean.hra.m2`.
```{r}
carnivores$mean.hra.m2 %>% anyNA()
herbivores$mean.hra.m2 %>% anyNA()
```
On average, herbivores have a larger `mean.hra.m2` than carnivores.
```{r}
herbivores$mean.hra.m2 %>% mean(na.rm = T) > carnivores$mean.hra.m2 %>% mean(na.rm = T)
```


**9. Make a new dataframe `deer` that is limited to the mean mass, log10 mass, family, genus, and species of deer in the database. The family for deer is cervidae. Arrange the data in descending order by log10 mass. Which is the largest deer? What is its common name?**  
The largest deer is Alces alces.
```{r}
deer <- homerange %>% filter(family == "cervidae") %>% 
  select(family, genus, species, mean.mass.g, log10.mass) %>% 
  arrange(desc(log10.mass))

deer %>% head(1)
```
The common name of the largest deer species is moose.
```{r}
homerange %>% filter(species == "alces") %>% select(common.name)
```


**10. As measured by the data, which snake species has the smallest homerange? Show all of your work, please. Look this species up online and tell me about it!** **Snake is found in taxon column**    
The namaqua dwarf adder has the smallest homerange among the snake species. The namaqua dwarf adder is found in coastal regions near the border between Namibia and South Africa and is one of the smallest vipers in the world.
```{r}
snakes <- homerange %>% filter(taxon == 'snakes') %>% 
  select(common.name, mean.hra.m2) %>% 
  arrange(mean.hra.m2)

snakes %>% head(1)
```


## Push your final code to GitHub!
Please be sure that you check the `keep md` file in the knit preferences.   